---
title: "National Weather Service Storm Data Analysis : Peer Graded Assignment - Reproducible Research"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
```


## Data Processing

Data processing was done following the next principles:

 1. Download file: Check if storm Datafile Exists if not, download. File name "StormData.csv.bz2"
 2. Read File: Read file
 3. Select variables to use using DPLYR (BGN_DATE, EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
 4. Change date format, BGN_DATE--- to as.Date()
 5. Filter Date using DPLYR ( >+ 1996-01-01) ---- All events are recorded from this date
 6.  Use Exponentials on CROP and PROPERTY damage.
 7. Group Data by event type, count frequency, sum fatalities, sim injuries, sum crop damage, sum prop damage and arrange by frequency
 8. Creat a column called TOTALMONETARYDAMAGE which includes the some of the CROP AND PROPERTY damage
 9. Upper event types to: Extract Summaries and Records
 10. Read.CSV with official event na,es, UPPER 
 11. Create Column with Boolean = TRUE if EVTYPE matches
 12. Looped EVTYPE with OFFICIAL EVENTS NAMES creating a new Column that matches
 13. Group data frama  arranging by Fatalities(desc)
 
Due to the high amount of work I am using a sample of the data (EVTYPE) names that matched with the OFFICIAL NAME FILE. With the sample I cover:
  
  * 86.19% of total Fatalities
  * 86.44% of total Injuries
  * 77.55% of total Crop Damage
  * 63.67% of total Property Damage
  * 64.87% of total Monetary Damage (Crop + Property)
```{r data_processing, include=FALSE, cache=TRUE}

download_file <- function(){
  url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
  if(!file.exists("StormData.csv.bz2")){
    
    file <- download.file(url, destfile = "./StormData.csv.bz2", method = "curl")
    return(file)
  }
}

download_file()
read_file <- function(){
  
    df <- read.csv(bzfile("StormData.csv.bz2"), strip.white=TRUE)
    df
}

df <- read_file()

df_variables <- select(df,BGN_DATE, EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)

df_variables$BGN_DATE <- as.character(df_variables$BGN_DATE)
df_variables$BGN_DATE <- as.Date(df_variables$BGN_DATE,format =  "%m/%d/%Y")

df_dated <- filter(df_variables, BGN_DATE >= "1996-01-01")



######################
############### Exponetials

K <- 1000
M <-1000000
B <- 1000000000
zero <-  10
blanK <- 0



df_exponent_crop <- mutate(df_dated, cropValue = ifelse(CROPDMGEXP == "K",CROPDMG * K,
                                                  ifelse(CROPDMGEXP == "M",CROPDMG * M,
                                                  ifelse(CROPDMGEXP == "B",CROPDMG * B,     
                                                  ifelse(CROPDMGEXP == "0",CROPDMG * zero,0    
                                                          )))))

df_exponent_crop_property <- mutate(df_exponent_crop, propValue = ifelse(PROPDMGEXP == "K",PROPDMG * K,
                                                                 ifelse(PROPDMGEXP == "M",PROPDMG * M,
                                                                  ifelse(PROPDMGEXP == "B",PROPDMG * B,     
                                                                  ifelse(PROPDMGEXP == "0",PROPDMG * zero,0    
                                                                               )))))

############## Need to arrange PROPDMGEXP CROPDMGEXP and then sum
#####group data
# 
df_grouped <- group_by(df_exponent_crop_property, EVTYPE) %>% summarize(freq = length(EVTYPE),
                                                       FATALITIES = sum(FATALITIES),
                                                       INJURIES = sum(INJURIES),
                                                       TotalCropDamage = sum(cropValue),
                                                       TotalPropDamage = sum(propValue)
                                                       

                                                       ) %>% arrange(desc(freq)) %>% mutate(TotalMonetaryDamage = TotalCropDamage + TotalPropDamage)




#################
############### get event type

##### clean summary

df_grouped$EVTYPE <- toupper(df_grouped$EVTYPE)

summary_index <- grepl("SUMMARY.+",df_grouped$EVTYPE)
NOT_summary_names <-  df_grouped$EVTYPE[!summary_index]
df_grouped <- filter(df_grouped, EVTYPE %in% NOT_summary_names)


summary_index <- grepl("RECORD+.",df_grouped$EVTYPE)
NOT_summary_names <-  df_grouped$EVTYPE[!summary_index]
df_grouped <- filter(df_grouped, EVTYPE %in% NOT_summary_names)



##########
########## analyize event names

read_events <- read.csv("event_names.csv", header = FALSE, strip.white=TRUE)
read_events <- toupper(read_events$V1)
names_event <- df_grouped$EVTYPE

 events_data_index <- which(read_events %in% names_event)
 events_in_data <- read_events[events_data_index]
 events_NOT_in_data <- read_events[-events_data_index]
############# Get events




# names_splitted <- unlist(strsplit(names_event," ", fixed = TRUE))
# names_splitted <- unlist(strsplit(names_event," ", fixed = TRUE))
# df_names <- arrange(as.data.frame(table(names_splitted)), desc(Freq))

df_grouped$EVTYPE <- as.factor(df_grouped$EVTYPE)

# df_grouped_types <- mutate(df_grouped, EventTypeCons = ifelse( EVTYPE %in% names[grep("*HAIL.*",df_grouped$EVTYPE)], "HAIL", "OTHER"),
#                            ifelse()
#                                                  )

df_grouped_types <- mutate(df_grouped, EventTypeCons_boolean = EVTYPE  %in%  read_events) 

i <- 1
df_grouped_types$EventTypeCons <- c(1:length(df_grouped_types$EVTYPE))
for(x in df_grouped_types$EVTYPE){
  # index <- which(read_events %in% names_event[i])
  # my_list[i] <- read_events[index]
  # 
  
  if(names_event[i] %in% read_events == TRUE ){
    index <- which(read_events %in% names_event[i])
    df_grouped_types$EventTypeCons[i] <- read_events[[index]]
  }
  else{
     df_grouped_types$EventTypeCons[i] <- "NO NAME"
     
   }
  i <- i + 1
}
  
df_last <- group_by(df_grouped_types,EventTypeCons)  %>% summarize(
                                                                  FATALITIES = sum(FATALITIES),
                                                                   INJURIES = sum(INJURIES),
                                                                   TotalCropDamage = sum(TotalCropDamage),
                                                                   TotalPropDamage = sum(TotalPropDamage),
                                                                   TotalMonetaryDamage = sum(TotalMonetaryDamage)
                                                                   
                                                                   ) %>% arrange(desc(FATALITIES)) %>% filter(EventTypeCons != "NO NAME")  
  


```



```{r analyze, include=TRUE,echo = TRUE,cache=FALSE}

dim(df)

```